---
title: "Thoughts on the Bechdel Test Module (day 18; Fall 2016"
author: "Brian Danielak"
output: 
  html_notebook:
    toc: true
    number_section: true
---

I feel like this lesson would have gone smoother if we let students know a few things.

# Working with building blocks is exactly what happens every time we need to learn a new package or a new API. 

Professionals do it every day. And packages/APIs may contain useful functions, but they also *might not*. So, deciding whether to use a piece is just as important as deciding *not* to use a piece ("`webtext.paras()` isn't what I want, because it returns an object with too much nestedness. I just want a 1 x n array-like thing where each row is a line of script."

# We should consider teaching them a basic grammar of data analysis

In R, the [`dplyr` package](https://cran.r-project.org/web/packages/dplyr/README.html) provides a wonderfully consistent grammar and interface for data operations. And there are just [a handful of elemental operations](http://r4ds.had.co.nz/transform.html) that get most data tasks accomplished. Below, I'll detail those core operations. Together, they form a shared grammar for talking about manipulations.

```{r, echo=TRUE, eval=TRUE}
library(datasets)
library(dplyr)
library(magrittr)
```


## Grab only certain rows using `filter`. 

```{r}
mtcars %>% filter(wt > 2.000) # get cars whose weight is greater than 2.000
```

## Sort rows by a value using `arrange`. 

For example, arrange all of these cars by their number of engine cylinders, and within that, sort them by their city mileage values
    
```{r}
mtcars %>% arrange(mpg, cyl)
```

## Select a subset of columns using `select`
    
```{r}
mtcars %>% select(mpg, cyl)
```

## Add new variables using `mutate`

Say we wanted to create a "relative efficiency index"

$$\textrm{relative efficiency} = \frac{\textrm{efficiency of a given car}}{\textrm{efficiency of the most efficient car in the dataset}} \times{} 100\%$$

We can do that, *and* we can compose operations to select just the relative efficiency column.

```{r}
mtcars %>% 
  mutate(car_name = rownames(mtcars)) %>% 
  mutate(relative_efficiency = mpg / max(mpg)) %>% 
  arrange(desc(relative_efficiency)) %>%  # arrange by relative efficiency in descending order 
  select(car_name, relative_efficiency, mpg)
```

## Group data and aggregate using `group_by` and `summarize`

Say we wanted to know the average mpg for each type of engine cylinder configuration.

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(average_mileage = mean(mpg))
```


# Closing Thoughts

There's a strong movement in the R community to build a common grammar for data operations. The community calls this the [tidyverse](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/): a collection of packages built around elemental grammars for data cleaning, transformation, visualization, and modeling.

- [tidy data]() means one row per observation
- 

At the end of class, Tony (who had been observing and asking me questions about the solutions throughout) sighed and said "That's very complicated."
